---
title: "Project Proposal"
author: "Saniya Bekova"
format: html
editor: source
---

## Data Sources

```{r load necessary libraries}
#| message: false
#load necessary libraries
library(dplyr)
library(tidyr)
```

```{r load data}
#| message: false
#| warning: false
#load data
education_data <- read.csv('/Users/saniyabekova/Desktop/UMASS/Fall 2024/Machine Learning/Final Project/data/SDG_Sept2024_.csv')  # Dataset 1: Educational indicators
tuesdata <- tidytuesdayR::tt_load('2024-09-24') 
imo_data <- tuesdata$country_results_df # Dataset 2: IMO competition results

```

1.  **UNESCO Institute for Statistics (UIS)**: Educational indicators data collected from <https://sdg4-data.uis.unesco.org/>. This data provides comprehensive educational metrics for various countries across multiple years and was last updated in September 2024. Data from `r min(education_data$Year)` to `r max(education_data$Year)`

2.  **TidyTuesday GitHub Repository**: Data related to the International Mathematical Olympiad (IMO) collected from <https://github.com/rfordatascience/tidytuesday/blob/master/data/2024/2024-09-24/readme.md>. The IMO data tracks country-level performance, including scores, medals, and rankings, and was also updated in September 2024. Data from `r min(imo_data$year)` to `r max(imo_data$year)`

## Units of Analysis

-   **Countries**: Each country represents a unit of analysis in this dataset, with attributes related to educational performance (such as completion rates, expenditure) and their success in the IMO (average score, team size, medals won).

-   **Time**: The dataset spans multiple years, allowing for the analysis of trends over time in both education indicators and IMO performance.

## Topic Description

This project aims to explore how a country’s education system impacts its performance in the **International Mathematical Olympiad (IMO)**. I am particularly interested in studying the relationship between **government spending on education, primary and secondary education completion rates, literacy rates** and a country's success in the IMO.

## Why This Topic?

This topic interests me because I want to understand how a country’s investment in education and the quality of its education system influence its ability to succeed in international academic competitions like the IMO. By exploring these relationships, I hope to identify the factors that most strongly contribute to winning medals or achieving high scores in the IMO.

## Expectations

I expect to find a **positive relationship** between a country’s investment in education and its success in the IMO. Specifically:

1.  Countries that invest more in education (as a percentage of GDP) will likely perform better in the IMO.
2.  Countries with higher **primary and secondary education completion rates** will have stronger IMO teams.
3.  **Higher literacy rates** will likely correlate with better overall performance in the IMO.

However, I also expect exceptions, such as smaller or less wealthy countries that may excel due to targeted investments in math education or specialized training programs.

```{r get inidcators of education data}
#get indicators of educational data from one column to multiple
primary_edu_completion_rate <- filter(education_data, Indicator.Name == "Completion rate, primary education, both sexes (%)") %>%
  select(Year, Country, Value)

lower_sec_edu_completion_rate <- filter(education_data, Indicator.Name == "Completion rate, lower secondary education, both sexes (%)") %>%
  select(Year, Country, Value)

upper_sec_edu_completion_rate <- filter(education_data, Indicator.Name == "Completion rate, upper secondary education, both sexes (%)") %>%
  select(Year, Country, Value)

gross_enr_ratio_for_tertirary_edu <- filter(education_data, Indicator.Name == "Gross enrolment ratio for tertiary education, both sexes (%)") %>%
  select(Year, Country, Value)

literacy_rate <-filter(education_data, Indicator.Name == "Youth literacy rate, population 15-24 years, both sexes (%)") %>%
  select(Year, Country, Value)

gov_expen_as_perc_of_GPP <- filter(education_data, Indicator.Name == "Government expenditure on education as a percentage of GDP (%)") %>%
  select(Year, Country, Value)

```

```{r merge indicators of education data}
# Sequentially merge indicators of education data

merged_education_data <- merge(primary_edu_completion_rate, lower_sec_edu_completion_rate, by = c("Country", "Year"), all = TRUE,
suffixes = c("_primary_edu_completion_rate", "_lower_sec_edu_completion_rate"))

merged_education_data <- merge(merged_education_data, upper_sec_edu_completion_rate, by = c("Country", "Year"), all = TRUE,
suffixes = c("", "_upper_sec_edu_completion_rate"))

merged_education_data <- merge(merged_education_data, gross_enr_ratio_for_tertirary_edu, by = c("Country", "Year"), all = TRUE,
suffixes = c("", "_gross_enr_ratio_for_tertirary_edu"))


merged_education_data <- merge(merged_education_data, literacy_rate, by = c("Country", "Year"), all = TRUE,
suffixes = c("", "_literacy_rate"))


merged_education_data <- merge(merged_education_data, gov_expen_as_perc_of_GPP, by = c("Country", "Year"), all = TRUE,
suffixes = c("", "_gov_expen_as_perc_of_GPP"))

```
## Summary
```{r summary of merged education data}
library(tidyverse)
library(readxl)
library(haven) #for loading other datafiles (SAS, STATA, SPSS, etc.)
library(stringr)
library(lubridate)
#getting summary of merged education data
summary(merged_education_data)
write_csv(merged_education_data, "data/education_data.csv")
```

```{r summary of imo_data}
#getting summary of imo_data
summary(imo_data)
write_csv(imo_data, "data/imo_data.csv")
```
